---
title: "pan_cancer_analysis"
author: "Niklas Rindtorff"
date: "5/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
```

#Introduction 
#Load packages

```{r}
library(tidyverse)
library(stringr)

```

#Load data
```{r}
load("crxg.Rdata")
```


#Correlation pan-cancer

I show the null-distribution of drug correlations for all drugs and cell lines available.

```{r}
crxgr_pan <- crxg %>% dplyr::select(drug_id, cosmic_id, rob_z) %>%
  filter(!is.na(drug_id), !is.na(cosmic_id)) %>%
  spread(drug_id, rob_z) %>% 
  dplyr::select(-cosmic_id)%>%
  corrr::correlate(method = "spearman", use = "pairwise.complete.obs") %>%
  corrr::stretch()

crxgr_pan %>%
  drop_na() %>%
  ggplot(aes(r)) + 
  geom_density() + 
  theme_classic()
```

```{r}
crxg_map <- crxg %>% dplyr::select(drug_id, cosmic_id, ln_ic50) %>%
  filter(!is.na(drug_id), !is.na(cosmic_id)) %>% 
  group_by(drug_id) %>% 
  mutate(med_ln_ic50 = median(ln_ic50, na.rm = TRUE)) %>% 
  mutate(norm_ln_ic50 = ln_ic50 - med_ln_ic50) %>%
  ungroup() %>% 
  dplyr::select(drug_id, cosmic_id, norm_ln_ic50) %>%
  spread(drug_id, norm_ln_ic50) %>% 
  column_to_rownames("cosmic_id") %>% 
  #drop_na() %>%
  as.data.frame() 

crxg_map %>%
  naniar::vis_miss()
```

Without dropping NAs the dimensions are: 

```{r}
crxg_map %>% dim()
```

With dropping NAs the dimensions are: 

```{r}
crxg_map %>%
  drop_na() %>% dim()
```


```{r}
anno_row  = crxg %>% 
  drop_na() %>%
                  dplyr::select(cosmic_id, cancer_type ) %>% 
                  distinct() %>%
  mutate(gi_cancer = if_else(cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD"), "GI", "other")) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  dplyr::select(-cancer_type) %>%
                  column_to_rownames("cosmic_id")

anno_col  = crxg %>% 
  drop_na() %>%
                  dplyr::select(drug_id, target_pathway) %>% 
                  distinct() %>%
  as.data.frame() %>%
  remove_rownames() %>%
                  column_to_rownames("drug_id")


crxg_map %>%
  drop_na() %>%
  pheatmap::pheatmap(na_col = "grey", scale = "column",
                     #annotation_col = anno_col, 
                     annotation_row = anno_row, 
                     show_rownames = FALSE, show_colnames = FALSE,
                     cutree_rows = 5, 
                     filename = "crxg.pdf", width = 9, height = 9)
                     
```

Idea: only show drugs with 6sigma activity
I now filter all GI lines and plot the compound responses for these cases. 

```{r}
crxg_map_gi <- crxg %>% 
  filter(cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>%
  dplyr::select(drug_id, cosmic_id, ln_ic50) %>%
  filter(!is.na(drug_id), !is.na(cosmic_id)) %>% 
  
  group_by(drug_id) %>% 
  mutate(med_ln_ic50 = median(ln_ic50, na.rm = TRUE)) %>% 
  mutate(norm_ln_ic50 = ln_ic50 - med_ln_ic50) %>%
  ungroup() %>% 
  dplyr::select(drug_id, cosmic_id, norm_ln_ic50) %>%
  spread(drug_id, norm_ln_ic50) %>% 
  column_to_rownames("cosmic_id") %>% 
  #drop_na() %>%
  as.data.frame() 

crxg_map_gi %>%
  #drop_na() %>%
  naniar::vis_miss()
```

```{r}
crxg %>% 
  filter(cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>% 
  group_by(drug_id, drug_name) %>% 
  summarise(var = var(ln_ic50),
            var_test = var(rob_center)) %>% View
  ggplot(aes(reorder(drug_name, - var), var)) + 
  geom_point()
```

```{r}
df <- crxg %>% 
  filter(cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>% 
  group_by(drug_name) %>%
  mutate(var = var(ln_ic50)) %>%
  left_join(., 
            crxg %>% 
  filter(cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>% 
  group_by(drug_name) %>%
  do(shapiro.test(.$ln_ic50) %>% broom::tidy()) %>% 
  mutate(p.adj = p.adjust(p.value, method = "BH"))) %>% 
  mutate(l.p.adj = -log(p.adj)) %>% 
  ungroup() %>%
  mutate(doi = if_else(l.p.adj > 19, TRUE, FALSE)) %>%
  dplyr::select(drug_name, var, l.p.adj, doi) %>% 
  distinct()

df %>%
  #ggplot(aes(reorder(drug_name, - p.adj), p.adj)) + 
  ggplot(aes(var, l.p.adj)) + 
  geom_point() + 
  ggrepel::geom_text_repel(data = df %>% filter(doi == TRUE), aes(label = drug_name)) +
  geom_hline(yintercept = 5) +
  #scale_y_log10() + 
  theme_classic() + 
  geom_rug() +
  ggsave("non_normal_drugs.pdf", width = 4, height = 4)
```

```{r}
df %>% 
  filter(l.p.adj >5) %>% View
```

```{r}
abnormal_drugs <- df %>% 
  filter(l.p.adj >5) %>% .$drug_name
```

filter only the abnormal drugs and 

```{r}
crxgr_pan %>%
  
  drop_na() %>%
  ggplot(aes(r)) + 
  geom_density() + 
  theme_classic()
```



```{r}
load("~/combis/crxgf.Rdata")
```

```{r}
crxgf %>% 
  #filter(!is.na(drug_id), !is.na(cosmic_id), cancer_type %in% keep_types) %>%
  #filter(cancer_type == "COAD/READ") %>% 
  filter(!is.na(drug_id), !is.na(cosmic_id), cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>%
  filter(drug_name %in% abnormal_drugs) %>%
  dplyr::select(drug_id, cosmic_id, rob_z) %>%
    spread(drug_id, rob_z) %>% 
    dplyr::select(-cosmic_id)%>%
    cor(method = "spearman", use = "pairwise.complete.obs") %>% 
  pheatmap::pheatmap(filename = "gi_cancer_cor_abnormal.pdf",
                     width = 10, height = 10)

crxgf %>% 
  #filter(!is.na(drug_id), !is.na(cosmic_id), cancer_type %in% keep_types) %>%
  #filter(cancer_type == "COAD/READ") %>% 
  filter(!is.na(drug_id), !is.na(cosmic_id), cancer_type %in% c("COAD/READ", "ESCA", "PAAD", "STAD")) %>%
  filter(drug_name %in% abnormal_drugs) %>%
  dplyr::select(drug_id, cosmic_id, rob_z) %>%
    spread(drug_id, rob_z) %>% 
    dplyr::select(-cosmic_id)%>%
    cor(method = "spearman", use = "pairwise.complete.obs") %>% View
```

Prepare a heatmap of differential cell line response

```{r, eval = FALSE}
anno_col  = crxg %>% 
  drop_na() %>%
                  dplyr::select(drug_id, target_pathway) %>% 
                  distinct() %>%
  as.data.frame() %>%
  remove_rownames() %>%
                  column_to_rownames("drug_id")


crxg_map_gi %>%
  drop_na() %>%
  pheatmap::pheatmap(na_col = "grey", scale = "none",
                     #annotation_col = anno_col, 
                     annotation_row = anno_row, 
                     show_rownames = FALSE, show_colnames = FALSE,
                     cutree_rows = 5, 
                     filename = "crxg_gi.pdf", width = 9, height = 9)
```

